Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-07-31 paper/paper.tex
Fix broken citations and minor fixes

\def\journal{Neuroinformatics}
\def\DOI{}
\def\articleType{Technology Report}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Gorur-Shandilya, Hoyland \& Marder} 
\def\Authors{Srinivas Gorur-Shandilya\,$^{1*\dagger}$, Alec Hoyland\,$^{1\dagger}$, and Eve Marder$^{1}$}
\def\Address{$^{1}$Volen Center for Complex Systems and Biology Department \\ Brandeis University \\ Waltham, MA 02453 \\ USA
\vspace{0.25cm}

\vspace{0.5em}\tiny{$\dagger$ These authors have equally contributed to this article.}
}
\def\corrAuthor{Srinivas Gorur-Shandilya }
%\def\corrAddress{Marder Lab, Brandeis University, Biology Department and Volen Center for Complex Systems, Waltham, MA, USA}
\def\corrAddress{Volen Center for Complex Systems, Brandeis University, Waltham MA 02453}
\def\corrEmail{\texttt{srinivasgs@brandeis.edu}}

@@

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an electrically excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of neuroscience, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, with several parameters. Second, many or all equations in these models can be strongly coupled through variables like the membrane potential.  In multi-compartment models of spatially extended neurons, membrane potentials in every compartment are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like "C++", "FORTRAN" or "hoc" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia" or "MATLAB".

@@

\paragraph{New mechanisms require new \texttt{C++} code.} Adding new network components requires writing new "C++" code. Adding a new conductance in the Hodgkin-Huxley formalism (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, dayanTheoreticalNeuroscience2001}), requires creating a new "C++" header file, though this is generally trivial. Implementing a new integration scheme or component type requires much more in-depth knowledge of the underlying "C++" core code.

@@

The axon of the giant squid contains contains a fast inactivating sodium conductance ("NaV"), a slower non-inactivating potassium conductance ("Kd"), and a passive leak current. Seminal work by Hodgkin and Huxley showed that depolarizations of the membrane, for example, by injecting current, could lead to an activation of the voltage-sensitive "NaV" channels, which led to a run away depolarization that was terminated by the inactivation of "NaV" channels and the activation of "Kd" channels that repolarized the membrane (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952}). As one of the simplest models of excitable neural membranes, the Hodgkin-Huxley model often serves as a the first model introduced in pedagogical literature (\cite{dayanTheoreticalNeuroscience2001, sterrattPrinciplesComputationalModelling2011, trappenbergFundamentalsComputationalNeuroscience2010}).

@@

Voltage clamping is a experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (cite D\&A). Under this paradigm, the membrane voltage is "clamped" or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized (\cite{coleIonicCurrentMeasurements1960, coleIonsPotentialsNerve1955, hodgkinEffectSodiumIons1949, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, turrigianoSelectiveRegulationCurrent1995}).

@@

- reproducibility crisis in computational science 
- unlike experiemntal research, where reproducability can arise due to deep reasons, reasons for computational research irreproducability are mostly trivial: 
- multiple sources: software being used is closed-source, researcher's code is not made available; obscure architecture makes improper use possible. 
- how xolotl conributes to reoridubile science: (1) freely available, code for this paper available too for others to inspect and replicate (2) intutitive syntax, solid documentation. (3) architectural choices that ensure accountability: object hashing, parameter hashing possible.

>>>>

\def\journal{Neuroinformatics}
\def\DOI{}
\def\articleType{Technology Report}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Gorur-Shandilya, Hoyland \& Marder} 
\def\Authors{Srinivas Gorur-Shandilya\,$^{1*\dagger}$, Alec Hoyland\,$^{1\dagger}$, and Eve Marder$^{1}$}
\def\Address{$^{1}$Volen National Center for Complex Systems and Biology Department \\ Brandeis University \\ Waltham, MA 02453 \\ USA
\vspace{0.25cm}

\vspace{0.5em}\tiny{$\dagger$ These authors have equally contributed to this article.}
}
\def\corrAuthor{Srinivas Gorur-Shandilya }
%\def\corrAddress{Marder Lab, Brandeis University, Biology Department and Volen Center for Complex Systems, Waltham, MA, USA}
\def\corrAddress{Volen National Center for Complex Systems, Brandeis University, Waltham MA 02453}
\def\corrEmail{\texttt{srinivasgs@brandeis.edu}}

@@

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an electrically excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952a}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of neuroscience, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, with several parameters. Second, many or all equations in these models can be strongly coupled through variables like the membrane potential.  In multi-compartment models of spatially extended neurons, membrane potentials in every compartment are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like "C++", "FORTRAN" or "hoc" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia" or "MATLAB".

@@

\paragraph{New mechanisms require new \texttt{C++} code.} Adding new network components requires writing new "C++" code. Adding a new conductance in the Hodgkin-Huxley formalism (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952a, hodgkinQuantitativeDescriptionMembrane1952a, dayanTheoreticalNeuroscience2001}), requires creating a new "C++" header file, though this is generally trivial. Implementing a new integration scheme or component type requires much more in-depth knowledge of the underlying "C++" core code.

@@

The axon of the giant squid contains contains a fast inactivating sodium conductance ("NaV"), a slower non-inactivating potassium conductance ("Kd"), and a passive leak current. Seminal work by Hodgkin and Huxley showed that depolarizations of the membrane, for example, by injecting current, could lead to an activation of the voltage-sensitive "NaV" channels, which led to a run away depolarization that was terminated by the inactivation of "NaV" channels and the activation of "Kd" channels that repolarized the membrane (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952a}). As one of the simplest models of excitable neural membranes, the Hodgkin-Huxley model often serves as a the first model introduced in pedagogical literature (\cite{dayanTheoreticalNeuroscience2001, sterrattPrinciplesComputationalModelling2011, trappenbergFundamentalsComputationalNeuroscience2010}).

@@

Voltage clamping is a experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (cite D\&A). Under this paradigm, the membrane voltage is "clamped" or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized (\cite{coleIonicCurrentMeasurements1960, coleIonsPotentialsNerve1955, hodgkinEffectSodiumIons1949, hodgkinMeasurementCurrentvoltageRelations1952a, hodgkinQuantitativeDescriptionMembrane1952a, turrigianoSelectiveRegulationCurrent1995}).

@@

- reproducibility crisis in computational science \\
- unlike experimental research, where reproducibility can arise due to deep reasons \\
- reasons for computational research irreproducibility are mostly trivial:  \\
- multiple sources: software being used is closed-source, researcher's code is not made available; obscure architecture makes improper use possible.  \\
- how xolotl contributes to reproducible science: (1) freely available, code for this paper available too for others to inspect and replicate (2) intuitive syntax, solid documentation. (3) architectural choices that ensure accountability: object hashing, parameter hashing possible.
